Concurrent Programming with CXL

### 2020-06-07 book/paper.tex
Starting on Split Binary Semaphores

What we want is a special kind of lock that allows either one writer
or one or more readers to be in the critical section.  This is called
a \emph{reader/writer lock}~\cite{courtois71}.
We will explore various ways of implementing reader/writer locks in
this chapter and future ones.

@@

>>>>

What we want is a special kind of lock that allows either one writer
or one or more readers to be in the critical section.  This is called
a \emph{reader/writer lock}~\cite{courtois71}.
[RVR: or was it David Parnas who first introduced them?]
We will explore various ways of implementing reader/writer locks in
this chapter and future ones.

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbinsem.cxl}
\end{code}
\caption{Reader/Writer Lock implementation using Split Binary Semaphores.}
\label{fig:RWsplitsema}
\end{figure}

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical regions} (CCRs).
All critical regions are in fact conditional.  Even the basic one
has the condition that a process can only enter if no other process
is in the critical region.  Reader/writer locks refined that, loosening
the restriction for improved performance.  In the case of bounded buffers,
there were two CCRs.  First, a CCR
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCR that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

CCRs are easy to implement using busy waiting:
\begin{code}
\begin{verbatim}
let blocked = True:
    while blocked:
        call lock();
        if application-condition-holds:
            blocked = False;
        else:
            call unlock():
@cr: application-code;
call unlock():
\end{verbatim}
\end{code}

We have seen solutions to the reader/writer problem and the producer/consumer
problem that do no busy-wait, but they are quite different.  It would be nice
to have a general technique for all CCR problems.
The ``Split Binary Semaphore'' is such a technique, originally proposed by
Tony Hoare and popularized by Dijkstra.
